```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(Hmisc)
```

## Read, process and summarize dataset

```{r, echo = TRUE}
activity_monitoring_data <- read.csv("data/activity.csv")
activity_monitoring_data <- mutate(activity_monitoring_data, 
                                   date = as.Date(date))
head(activity_monitoring_data)
str(activity_monitoring_data)
summary(activity_monitoring_data)
```

## Total number of steps per day

```{r, echo = TRUE}
total_number_of_steps_per_day <- with(activity_monitoring_data, 
                                      tapply(steps, date, sum, na.rm = TRUE))
plot(total_number_of_steps_per_day, 
     main = 'Total number of steps per day', 
     xlab = paste('Day ( starting on ',as.character(names(total_number_of_steps_per_day)[1]),')'), ylab = 'Number of steps', 
     type = 'h', lwd = 7, lend=1)
```

### Mean of total number of steps per day

```{r, echo = TRUE}
mean(total_number_of_steps_per_day, na.rm = TRUE)
```

### Median of total number of steps per day

```{r, echo = TRUE}
median(total_number_of_steps_per_day, na.rm = TRUE)
```

## Average daily activity pattern

```{r, echo = TRUE}
average_daily_activity_pattern <- with(activity_monitoring_data, 
                                      tapply(steps, interval, mean, na.rm = TRUE))
plot(names(average_daily_activity_pattern), average_daily_activity_pattern, 
     main = 'Average daily activity pattern', 
     xlab = 'Daily time interval [5 min]', ylab = 'Average daily number of steps', 
     type = 'l', lwd = 1)
```

### Interval with maximal average daily number of steps and corresponding value

```{r, echo = TRUE}
average_daily_activity_pattern[which.max(as.numeric(average_daily_activity_pattern))]
```

## Imputing missing values

### Total number of missing values in the dataset

```{r, echo = TRUE}
na_row_indices <- as.logical(rowSums(is.na(activity_monitoring_data)))
sum(na_row_indices)
```

### Filling in missing values by mean of total number of steps

```{r, echo = TRUE}
activity_monitoring_data_imputed <- mutate(activity_monitoring_data, steps = impute(steps, mean))
```

### Imputing strategy increased/creates some of the total number of steps per day

```{r, echo = TRUE}
total_number_of_steps_per_day_imputed <- with(activity_monitoring_data_imputed, 
                                      tapply(steps, date, sum, na.rm = TRUE))
plot(total_number_of_steps_per_day_imputed, 
     main = 'Total number of steps per day (imputed)', 
     xlab = paste('Day ( starting on ',as.character(names(total_number_of_steps_per_day_imputed)[1]),')'), ylab = 'Number of steps', 
     type = 'h', lwd = 7, lend=1)
```

### Imputing strategy increases mean and median slightly

```{r, echo = TRUE}
mean(total_number_of_steps_per_day_imputed, na.rm = TRUE)
median(total_number_of_steps_per_day_imputed, na.rm = TRUE)
```

## Activity patterns: weekdays vs. weekends

### Create new factor variable type.of.day with values "weekday" and "weekend"

```{r, echo = TRUE}
activity_monitoring_data <- mutate(activity_monitoring_data, type.of.day = "weekday")
weekend_indices <- weekdays(activity_monitoring_data$date) %in% c("Saturday", "Sunday")
activity_monitoring_data[weekend_indices, ]$type.of.day <- "weekend"
activity_monitoring_data <- mutate(activity_monitoring_data, type.of.day = as.factor(type.of.day))
```

### Average daily activity pattern: weekdays vs. weekends

```{r, echo = TRUE}
par(oma = c(0, 0, 2, 0))
par(mfrow = c(1, 2))
average_daily_activity_pattern_weekday <- with(activity_monitoring_data[activity_monitoring_data$type.of.day == "weekday", ], 
                                               tapply(steps, interval, mean, na.rm = TRUE))
average_daily_activity_pattern_weekend <- with(activity_monitoring_data[activity_monitoring_data$type.of.day == "weekend", ], 
                                               tapply(steps, interval, mean, na.rm = TRUE))
plot(names(average_daily_activity_pattern_weekday), average_daily_activity_pattern_weekday, 
     main = 'Weekdays', 
     xlab = 'Daily time interval [5 min]', ylab = 'Average daily number of steps', 
     type = 'l', lwd = 1, lty = 1, ylim = c(0, 250))
plot(names(average_daily_activity_pattern_weekend), average_daily_activity_pattern_weekend, 
     main = 'Weekends', 
     xlab = 'Daily time interval [5 min]', ylab = 'Average daily number of steps', 
     type = 'l', lwd = 1, lty = 1, ylim = c(0, 250))
title("Average daily activity pattern", outer=TRUE)
```
